Studying public transit API query logs\\ to get an indication of travel flows

### 2016-01-06 paper.tex
Fixes in Section 3.1

Results from preliminary analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
Three observations in particular prompted further analysis of the data in greater detail. 
First, there has been a steady increase in the number of queries made between 2012 and 2015. 
\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
A dinstinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively. 
Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred. This suggests that \todo{Pieter: what does this mean? The numbers appear rather small though. Ideally this should be reflected in figure 1}.

@@

Secondly, the daily and hourly distribution of queries appear to correspond with known commuting patterns. 
\cref{fig:average} depicts an break down of the queries that occur on an hourly basis for each day of the week.
Several trends can be observed in this chart. 
Morning and evening peak periods in particular, are clearly distinguishable.
Peak hours on weekdays are also distinctively different from those of weekends.
Similarly, the frequency of queries during the morning peak period on Fridays appear to be substantially lower than those that occur between Monday and Thursday.
The discrepancy maybe explained by a decrease in the number students who travel home from college over the weekends. \todo{Can we give a link to more reading material about this behaviour? This might be weird for US people? --> not sure if this is really case. We need to explain this in a clear way}.
The absence of a clear peak on Saturdays can be attributed to lack of journey to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.

Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
Figure 3 provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis. \todo{where people are heading between certain time intervals?}

@@

@@

Matching pairs of origin and destination (OD) stations must be aggregated prior to visualisation.
Several stages of data processing are required to arrive at this outcome.
As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualisation.
To overcome this challenge, an open-source reconciliator (https:// github.com/ irail/ stations) was employed to reduce spelling variations and correct identifiable errors.

>>>>

Results from preliminary analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
Three observations in particular prompted further analysis of the data in greater detail. 
First, there has been a steady increase in the number of queries made between 2012 and 2015. 
\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
A dinstinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively. 
Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.

@@

Secondly, the daily and hourly distribution of queries appear to correspond with known commuting patterns. 
\cref{fig:average} depicts a break down of the queries that occur, on average, on an hourly basis for each day of the week.
Several trends can be observed in this chart. 
Morning and evening peak periods in particular, are clearly distinguishable.
At noon, a small dent is noticable, which may be attributed to part time workers and meetings on location after or before noon.
Peak hours on weekdays are also distinctively different from those of weekends.
Similarly, the frequency of queries during the evening peak period on Fridays appear to be substantially higher than those that occur between Monday and Thursday.
The discrepancy maybe explained by students that travel home from their student homes over the weekends. 
The absence of a clear peak on Saturdays can be attributed to lack of journeys to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.

@@

Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
\cref{fig:dec24} provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.%\todo{where people are heading between certain time intervals?}

@@

Matching pairs of origin and destination (OD) stations must be aggregated prior to visualisation.
Several stages of data processing are required to arrive at this outcome.
As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
Queries made by automated user agents like search enginebots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualisation.
To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.
